#**Understanding Prosper Loan Data Set**
<p><small>*Uir&aacute; Caiado. April 7, 2015*</small></p>
------------------
```{r, include=FALSE, cache=TRUE}
#install needed packages
# install.packages('knitr',repos="http://cran.rstudio.com/")
# install.packages("tidyr")

#load libraries that I intend to use
library("plyr") 
library("tidyr")
library("dplyr")
library(ggplot2)
library("ggthemes")
library(gridExtra)
library(reshape2)

#load data
s_fname <- "~/git/ProsperLoan_EDA/prosperLoanData.csv"
d <- read.csv(s_fname)
```

##Abstract

I explore a data set about P2P loans. My goals with the study are understanding 
how this practice works, and what makes a loan to be bound to a specific risk 
score.

##Introduction

I have to tell the truth. Initially, the number of variables intimidated me. 
There are 113,937 rows in this data set and 81 columns (variables). It seems to 
come from Prosper website, a peer-to-peer [P2P](http://en.wikipedia.org/wiki/Peer-to-peer_lending)
lending marketplace.  I have already seen a brazilian website trying to do 
something similar in my country, but our "SEC" shut them off. Just financial 
companies can lend money here.

According to [wikipedia](http://en.wikipedia.org/wiki/Prosper_Marketplace#Loan_performance_prior_to_July_2009) ,
the SEC imposed that Prosper had to cease their operations  due to violations on 
[a piece of legislation](http://www.investopedia.com/terms/s/securitiesact1933.asp)
that ensures transparency and try to avoid fraudulent activities in the american
securities markets. The website took from November 2008 to July 2009 to resume 
their activities. 

Bellow you can see the effect of this time closed:

```{r Number of Loans by Year, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#creating new variables: year of origination
d$LoanOriginationDate <- as.Date(d$LoanOriginationDate,format = "%Y-%m-%d")
d$LoanOriginationDate.year=format(d$LoanOriginationDate, "%Y")

#plot the numb of loand by year of origination
ggplot(aes(x = LoanOriginationDate.year), data = d) +
  geom_histogram() + 
  ggtitle('Number of Loans by Year')
```

There are 6 variables in the data set just valid to loans originated after 2009 
and I believe that they were created to comply regulatory rules. In general, 
they are all related to how much interest have been paid and how risky each loan
were. Looking at other variables, I can see that they are pretty diverse but 
they are all related to these two factors: risk and return.

There are variable directly associated with risk measurements (what was the 
classification when the loan was took, the scores from Prosper), about how much 
was paid and how much was earned, meta data of the loan (the size in money and 
duration, when and where it was originated), about the profile of the borrower 
(employment status, his income, his debts, what he intend to do with the money),
how each loan was founded and so on.

In the next sections, I will explore 11 variables of the data set, as described
bellow:

* Term: The length of the loan expressed in months;
* LoanStatus: The current status of the loan, as Cancelled, Defaulted, 
  etc;
* ClosedDate: Closed date, when applicable;
* BorrowerRate: The Borrower's interest rate for the loan;
* ProsperRating (numeric): The  Risk Rating assigned at the time the listing
  was created. Varies from 0 (worst) to 7 (best);
* ListingCategory: The category of the listing that the borrower selected 
  when posting their listing;
* IsBorrowerHomeowner: If the Borrower is classified as a homeowner by the 
  Prosper Criteria;
* DebtToIncomeRatio: The debt to income ratio of the borrower at the time 
  the credit profile was pulle. This value is capped at 10.01;
* StatedMonthlyIncome: The monthly income the borrower stated at the time 
  the listing was created;
* LoanOriginalAmount: The origination amount of the loan;
* LoanOriginationDate: The date the loan was originated.

As P2P lending is something new for me, I would like to understand how it works 
in United States and I will use this database to achieve that.

##Analysis

###Univariate Section
<p>&nbsp;</p>
####Loan Meta Data: Amount Borrowed by Month of Loan Origination

In Brazil, where I live, I would expect the total loans rose at some months of 
the year. Usually at the beginning and the end of the year, when people travel, 
pay taxes and buy gifts. Let???s start looking at the amount of money borrowed by 
month in United States

```{r Money Borrowed By Month, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#creating new variables: month of origination
l_months <- c('Jan','Feb','Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep','Oct',
              'Nov', 'Dec')
d$LoanOriginationDate.month=format(d$LoanOriginationDate, "%b")
d$LoanOriginationDate.month <- factor(d$LoanOriginationDate.month, l_months)

#bar plot with the count of lons by month of origin
ggplot(aes(x = LoanOriginationDate.month, y = LoanOriginalAmount), data = d) +
  geom_bar(stat="identity") +
  ggtitle('Money Borrowed By Month')
```

As expected, the amount of loans have risen at first and last months of the year.

***

####Loan Meta Data: Amount Defaulted by Month of Loan Origination

If there are more money borrowed on these periods, maybe the number of 
loans  defaulted increases between them. To verify that, I have to use the 
ClosedDate  variable and filter just the defaulted and chargedoff loans. Let's 
see:

```{r Amount Defaulted By Month, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))
#creating new variable and set it as a factor
l_months <- c('Jan','Feb','Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep','Oct',
              'Nov', 'Dec')

d$ClosedDate <- as.Date(d$ClosedDate,format = "%Y-%m-%d")
d$ClosedDate.month=format(d$ClosedDate, "%b")
d$ClosedDate.year <- format(d$ClosedDate, "%Y")
d$ClosedDate.month <- factor(d$ClosedDate.month, l_months)
#plot histogram summing up LoanOriginalAmount and not just counting
ggplot(aes(x = ClosedDate.month, y = LoanOriginalAmount), 
       data = subset(d, (LoanStatus == 'Defaulted') |
                       (LoanStatus == 'Chargedoff'))) +
  geom_bar(stat="identity") +
  ggtitle('Amount Defaulted By Month')
```

Curriously the distribution of the amount borrowed by month looks like the 
distribution of amount defaulted/chargedoff by month. 

***

####Loan Meta Data: How Defaulted Loans are distributed

Well, I am not sure if this filtered data is relevant, let me see how many data 
points there are in each status. I will group all 'Past Due' buckets as the same
buckets and show off the amount of data on each status as percentage. As can be 
seem bellow, 11% of the data is related to defaulted or chargedoff loans:

```{r stats of loan status, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#create function to rename labels
rename_status <- function(s_status) {
  #return a string with renamed status passed
    if(substr(s_status, 0, 4)=='Past'){
      'Past Due Date'
    }else if(s_status == 'Chargedoff' | s_status =='Defaulted' ){
      'Chargedoff/Defaulted'
    }else{
      s_status
    }
}

#create a new variable: renamed loan status
d$LoanStatus.renamed <- apply(d['LoanStatus'],1,rename_status)
my_summ <- table(d$LoanStatus.renamed, useNA = 'ifany')

#normalize the count of status already renamed
round(my_summ/sum(my_summ)*100)/100

```

Thinking again, nobody would borrow money to pay back in the next couple of 
months, at least not in Brazil. Let's see how many loans were borrowed, 
percentually,  divided by duration:

```{r stats of terms, echo=FALSE, cache=TRUE}
#count different terms
my_summ <- table(d$Term)
round(my_summ/sum(my_summ)*100)/100
```

It explains why the distributions of Defaulted/Chargedoff loans and the amount 
borrowed by month looks like the same. As people take loans on regular durations,
12, 36 and 60 months, the loans will ever due to the same month in different years. 

***

####Loan Meta Data: Loan Original Amount and how it is distributed by Term

What about the median of the loan? First, let's see the quartiles of the value of
each loan:

```{r summary of LoanOriginalAmount, echo=FALSE, cache=TRUE}
#show quartiles of LoanOriginalAmount
summary(d$LoanOriginalAmount)
```

Wow...$ 1,000 to $35,000. It is a huge range. But it seems that the most of the 
loans are smaller, given that 50 % of the data set is up to $ 6500. 

Let's take a close look at the loan original amount splitted by duration. I had 
to convert the variable "Term" to factor to make the chart below:

```{r Box Plots of Amount of Loan by Term, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))
#set Term as a factor to show it not as a continuous number
d$Term.f <- as.factor(d$Term)

#box plot of the loan amount by term
ggplot(aes(x = Term.f, y = LoanOriginalAmount), data = d) +
  scale_color_brewer(type = 'div') +
  geom_boxplot(outlier.size = 0, color = "black") +
  ggtitle('Amount of Loan by Term')
```

It seems that the loan for each duration is something that I should not plot 
together. The distribution of each one is pretty different. 

***

####Loan Return: Take a Look at the Distribution of Borrower Rates

Well, and the interest rate? The distribution differs at each term? First, let's
see the histogram of the interest paid:

```{r Number of Loans by Interest, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#create summary of borrower rate
summary(d$BorrowerRate)

#plot a histograma of the borrower rate
ggplot(aes(x = BorrowerRate), data = d) +
  geom_histogram() + 
  ggtitle('Frequency of Interest Rate')
```

There is something different on the right tail. Let's reduce the bin width to 0.5%

```{r Number of Loans by Interest binwidth, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#plot a histograma of the borrower rate with modify binwidth
ggplot(aes(x = BorrowerRate), data = d) +
  geom_histogram(binwidth = 0.005) + 
  ggtitle('Frequency of Interest Rate (bin width =  0.5%)')
```

Definitely there is something different after 30% of annual interest rate. 

***

####Loan Return and Loan Meta Data: Investigating Outliers in the Interest Rates

Let's take a close look at the range between 30%-37%.

```{r Number of Loans by Interest binwidth 2, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#plot a histograma of the borrower rate with modify binwidth AND limiting 
#the values in X axis
ggplot(aes(x = BorrowerRate), data = d) +
  coord_cartesian(xlim=c(0.3, 0.37)) + 
  geom_histogram(binwidth = 1/1000) + 
  ggtitle('Frequency of Interest Rate (bin width =  0.5%)')

```

Now, more specifically in the range 31.5% and 32%. 

```{r, echo=FALSE, cache=TRUE}
#count borrower rate in an interval
table(subset(d, BorrowerRate>=0.316 & BorrowerRate<0.32)$BorrowerRate)
```

Curiously there is a huge concentration at 31.77% and 31.99%. Let me see when 
those loans were took.

```{r 3177chart, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#plot a histograma of the borrower rate with modify binwidth and with limits 
#in X axis
ggplot(aes(x = LoanOriginationDate),
       data = subset(d, BorrowerRate==0.3177 | BorrowerRate==0.3199)) +
  geom_histogram() + 
  ggtitle('Number of 31.77% and 31.99%\n interest rate Loan by Date')
```

Ok, I can see that there is no evidence of relationship with date. Maybe I find 
something related to the category of each term:  

```{r, echo=FALSE, cache=TRUE}
#count borrower rate in an interval
table(subset(d, BorrowerRate==0.3177 | BorrowerRate==0.3199)$Term)
```

Interesting, all loans with these interest rates have a term of 36 months. Let's
see how the rates are distributed by Term, considering all the data set.

```{r Frequency Polygons, , echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#frequency polygon by term
ggplot(aes(x = BorrowerRate), data = d) +
  geom_freqpoly(aes(color = Term.f)) + 
  scale_colour_brewer("Blues") +
  scale_x_continuous() +
  xlab('BorrowerRate') + 
  ylab('Count\n') +
  ggtitle('Count of loans by \nInterest Rate and Term\n')

```

The most of loans have a term of 36 months. It is curious to see that just loans 
with term of 36 months have a more dispersed distribution. However, it is 
difficult to see how the how loans with term of 12 months are distributed. 
Below are the Box plots.

```{r Box Plots of BorrowerRate by Term, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#box plot of BorrowerRate by Term without dots showing outliers
ggplot(aes(x = Term.f, y = BorrowerRate), data = d) +
  scale_color_brewer(type = 'div') +
  geom_boxplot(outlier.size = 0, color = "black") +
  ggtitle('Interest Rate by Term')
```

As previously observed on the frequency polygons , the box plot confirms that 
36-months loans are more dispersed. Additionally, I can notice that the median 
increases sharply from 12 to 36 months and then increases slightly to 60 months.
It makes sense, once that a long-term loan is much more unpredictable than a 12 
months loan.

***

####Loan Risk: Prosper Scores

Following my reasoning, if the borrower rate are dispersed, I would expected 
that the Risk Score also was dispersed, given that the interest rate should be a 
function of the risk score. First, let me see the quartiles of the Prosper 
Rating.

```{r, echo=FALSE, cache=TRUE}
#count and summarize ProsperRating
table(d$ProsperRating..numeric., useNA = 'ifany')
summary(d$ProsperRating..numeric., useNA = 'ifany')
```

As I can see, the risk scores are discrete and ordered. I believe that it is not
a issue, given that a average rate of 5.3, for instance, has a meaning. It tells
me that there is more data point on 5 than in 6. I also notice that there are 
many data point without a risk score. Below is a histogram just including those 
that has a score setted:

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#factorize the ProsperRating..numeric. variable 
d$ProsperRating..numeric. <- as.factor(d$ProsperRating..numeric.)

#histogram of ProsperRating
ggplot(aes(x = ProsperRating..numeric.), 
       data = subset(d,!is.na(ProsperRating..numeric.) )) +
  geom_histogram()
```

Well...They are not dispersed....and, thinking again, actually they shouldn't be. 
The risk score is discrete, there is no way to be dispersed. I would need verify
the distribution splitting the data by term, but I will check it later. 

***

####Borrower Profile: Borrower Monthly Income

Maybe the income can help me understand something:

```{r, echo=FALSE, cache=TRUE}
#summarize StatedMonthlyIncome and income range 
summary(d$StatedMonthlyIncome, useNA = 'ifany')
summary(d$IncomeRange, useNA = 'ifany')
```

75% of the incomes declared are less than or equal to $ 6,825 and the maximum 
income value is  $ 1,750,000. If I look at the income range, I can see that it 
was grouped just until $100,000. 

Let's look at data where the monthly income less than 100,000 and has a prosper 
score setted.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#histogram of income range filtered
ggplot(aes(x = StatedMonthlyIncome), 
       data = subset(d, d$StatedMonthlyIncome < 10000 &
                       !is.na(ProsperRating..numeric.))) +
  geom_histogram()
```

It doesn't add up much to my thoughts. I believe that these variables have to be 
related to another to be more meaningful. 

***

####Loan Meta Data: What Were The Reasons For The Loans ?

Now, let's see why people borrow money,considering just who has a prosper 
rating setted. 

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#histogram of listing category filtered
ggplot(aes(x = ListingCategory..numeric.), 
       data = subset(d, !is.na(ProsperRating..numeric.))) +
  geom_histogram() + 
  scale_y_continuous(breaks = seq(2000, 54000, 10000))

```

I need to rename the categories. Also, I am already seeing that there are a lot 
of categories that has less than 2,000 observations. I am going to rename those 
to 'Other'.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(14))

#create function to rename labels
rename_listingCat <- function(i_listing) {
  #list of new labels names
  l_names <- c('Not Available','Debt consolidation','Home Improvement',
               'Business', 'Personal Loans','Student Loan','Auto','Other',
               'Other','Other', 'Other','Other','Other','Household Expenses',
               'Other', 'Medical/Dental')
  
  #return a string with renamed status passed
  if (i_listing>15) {'Other'}
  else {l_names[i_listing+1]}
}

#create a new variable with renamed loan status
d$ListingCategory.renamed <- apply(d['ListingCategory..numeric.'],1,
                                   rename_listingCat)

#histogram of income range filtered
ggplot(aes(x = ListingCategory.renamed), 
       data = subset(d, !is.na(ProsperRating..numeric.))) +
  geom_histogram() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Much better. The loans to roll over existing debts are predominant. Let's see 
now how people were deep in debt when they took the loans using Prosper. 

***

####Borrower Profile: How Much of the Monhtly Income Is Committed to Loans ?

Given that the loans labeled as Debt Consolidation are most common, I would 
expect that the ratio debt o income be greater. Let's see how the ratio is
distributed.

```{r, echo=FALSE, cache=TRUE}
#quantiles for the DebtToIncomeRatio
summary(subset(d, !is.na(DebtToIncomeRatio))$DebtToIncomeRatio)
```

There are some outliers here. I am going to filter out the top 1% of debt/income
ratio in the next plot. Also, I will apply a log transformation here.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#filtered data, excluding the DI greater than 99% of the collection
filtered_d <- subset(d,!is.na(DebtToIncomeRatio))
exclude <- quantile(filtered_d$DebtToIncomeRatio, 0.99)

#histogram of DI and the same histogram with the levels of DI log-transformed
p1 <- ggplot(aes(x = DebtToIncomeRatio), 
       data = subset(filtered_d, DebtToIncomeRatio<exclude)) +
  geom_histogram(binwidth = 1/15)
  
p2 <- p1 + 
  scale_x_log10() + 
  xlab("log(DebtToIncomeRatio)")

#plot p1 and p2 into a unique figure
grid.arrange(p1, p2, ncol=1)

```

I am seeing that there is some kind of distribution here. The most of people 
have less than 30% of their income committed to debts. Also, I believe that 
there is no need to transform this variable in my future explorations.

***

####Wrapping Up the Univariate Section

So far, I have seen some evidences that people take more loans in specific months 
and the interest rate of these loans are strongly related to the term of the each
loan. There is something curious about loan with 37%-38% of interest rate and I 
have not understood why loan with term of 36 month are so much more dispersed 
than 60 months, although the median of interest rates on 60 month are slightly 
greater than 36 months. 

Also, I didn't find nothing so interesting about risk score and income level, two
variable that I was expecting more. On the other hand, maybe this variables have 
more to tell when related to others. I would like to understand better why loans
with term of 36-months are more dispersed than 60 months. Also I want to find 
out why risky loans are classified that way.


###Bivariate Section
<p>&nbsp;</p>

####Loan Return And Loan Risk: Daes The Risk Pays Off ?

Let's start looking at how risk score and borrower rate relate with each other. 
I'm expecting to find a strong relationship between them. After all, the interest
rate should reflect the risk of the loan. I am going to add some jitter on the 
chart and change the alpha to reduce the overplotting. I am also  excluding 
monthly income greater than $ 100,000.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#scatter plot of Borrower Rate by risk. I added jitter to reduce overplotting
ggplot(aes(y = BorrowerRate, x = ProsperRating..numeric.), 
       data = subset(d, d$StatedMonthlyIncome < 10000 &
                       !is.na(ProsperRating..numeric.))) +
  geom_point(alpha = 1/20, position = position_jitter( h = 0))
```

As the Prosper Rating is a discrete variable and the Borrower Rate variable  is 
continuous, the chart without the jitter can look like a bit cluttered. I can 
notice in the figure above that as better the risk score, more common are lower 
rates.

This chart also shows that loans with higher interest rate are much more common 
in loans  with higher risk (smaller scores). Let me see the correlation between 
these two variables.

```{r, echo=FALSE, cache=TRUE}
#correlation test
d$ProsperRating..numeric. <- as.integer(d$ProsperRating..numeric.)
with(subset(d, d$StatedMonthlyIncome < 10000 &
               !is.na(ProsperRating..numeric.)),
     cor.test(ProsperRating..numeric., BorrowerRate,
              method = 'pearson'))
```

As expected, there is a strong negative relationship between the two variables, 
confirming that smaller scores results in higher interest rates. As I said 
before, It was expected. 

***

####Borrower Profile And Loan Risk: Are The Wealthy People Less Risky?

What I would like to understand is WHY this loans were classified as risky. Let 
me see what is the relationship between risk score and monthly income. I am also
filtering out monthly income equal to 0 (zero).

```{r, echo=FALSE, cache=TRUE}
#function to be used to update my filtered data throughout the project
filter_data <- function(data){
  data_aux <- subset(data, data$StatedMonthlyIncome < 10000 &
               data$StatedMonthlyIncome > 0 &
               !is.na(ProsperRating..numeric.))
  return(data_aux) 
} 

```


```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#factorize prosper rating and filtering data again
d$ProsperRating..numeric.f <- factor(d$ProsperRating..numeric.)
filtered_d <- filter_data(d)

#box plot of income range splitted by risk
ggplot(aes(y = StatedMonthlyIncome, x =ProsperRating..numeric.f ), 
       data = filtered_d ) +
  geom_boxplot()

```

Looking at the chart above, It seems that there is some relationship between 
the two variables tested. The greater the monthly income, it is classified with 
a better risk profile more commonly . On the another hand, I am also seeing that
even people with high income are classified as high risk. The dispersion of the 
observations suggested that, the monthly income, alone, does not explain the 
Risk Score. Let's see the correlation test between the two variables.

```{r, echo=FALSE, cache=TRUE}
#filtering data and compute correlation between risk and income
filtered_d <- filter_data(d)
with(filtered_d, cor.test(ProsperRating..numeric., StatedMonthlyIncome,
                          method = 'pearson'))
```

The test shows that the correlation is significant, given that the p-value is 
lower than 0.05 and indicates that as higher the monthly income, better is the 
risk score, as I observed in the box plot. The correlation coefficient, around 
0.2, indicates that this relationship is not strong.

***

####Borrower Profile And Loan Risk: Are People in Debt Worst Classified ?

Maybe the correlation between the Prosper Rating and the Debt to Income ratio 
is stronger.

```{r, echo=FALSE, cache=TRUE}
#compute correlation between risk and DI
with(filtered_d , cor.test(ProsperRating..numeric., DebtToIncomeRatio,
                           method = 'pearson'))
```

This test also shows that the correlation is significant and indicates that the 
debt/income ratio and the Prosper Rating are negatively correlated. Said another 
way, the higher the debt/income ratio, worst is the Prosper Rating. But, again, the
correlation coefficient, around -0.1, is not strong. 

I wonder if there is any pair of variables, that is not Prosper Rating and 
Borrower Ratio, that presents a strong correlation between them.

***

####Loan Meta Data And Loan Risk: Loan Amount and Risk

Maybe there is something about the amount of the loan and Risk Score. I 
going to keep the same filters on the data that I used before and calculate the
mean loand amount on each risk category.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#bar plot with the average loan amount by risk category, what is calculated in 
#the layer stat_summary
ggplot(aes(y = LoanOriginalAmount, x =ProsperRating..numeric. ), 
       data = filtered_d ) + 
  stat_summary(fun.y="mean", geom="bar")  

```

It seems as some values of loans were allowed just for certain risk scores, but 
it does not add up to my understanding of Risk classification. 

***

####Loan Meta Data And Loan Risk: Listing Category on each Risk Score

Let me see how the loans for each risk score are divided by ListingCategory.

```{r, echo=FALSE, cache=TRUE, fig.width=9}
#set theme
theme_set(theme_minimal(17))

#grouping by prosper rating and listing category, then summarizing 
#listing category with counts
listing_groups <- group_by(filtered_d ,
                           ProsperRating..numeric.f,
                           ListingCategory.renamed)
d.risk_by_listing <- summarise(listing_groups,Count = n())

#scatter plot to show how numerous is each listing category by prosper rating
ggplot(aes(y = Count, 
           x = ProsperRating..numeric.f, 
           color = ListingCategory.renamed),
       data = d.risk_by_listing) +
  scale_color_economist() +
  geom_point(size = 4 , alpha = 0.7) +
  ggtitle("# of loans by prosperRating\n and Listing Category\n")


```

There are a lot of loans that were taken for Debt Consolidation on all Risk 
levels. If this listing category was predominant on all levels, it suggests to 
me that it is not a driver of risk level. Thus, I will not use this variable 
further on my analysis. 

***

####Borrower Profile and Return: Debts and Interest Rate

And how Debt to Income Ratio is related to Borrower Rate?

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#scatter plot of Borrower Rate by DI
ggplot(aes(x = BorrowerRate, y = DebtToIncomeRatio), 
       data = subset(d,  !is.na(ProsperRating..numeric.))) +
  geom_point(alpha = 0.5) +
  ggtitle("Distribution of Borrower Rate by Debt to Income category")
```

It is confuse. I am going to break up the Debt To Income Ratio into a
categorical one and plot the two variables again. The most of classes I have 
defined as the quartiles of the data.

```{r, echo=FALSE, cache=TRUE, fig.width=9}
#set theme
theme_set(theme_minimal(14))

#break up variable into buckets
d$DebtToIncomeRatio.bucket <- cut(d$DebtToIncomeRatio, 
                                  breaks = c(0,0.14,0.22,0.31,1, 2, 4, 6, 10.2))
#I have to filtered again my data set
filtered_d <- filter_data(d)

#box plot of interest rate by DI buckets
ggplot(aes(y = BorrowerRate, x = DebtToIncomeRatio.bucket), 
       data = subset(d,  !is.na(ProsperRating..numeric.))) +
  geom_boxplot() +
  ggtitle("Distribution of Borrower Rate by Debt to Income category")
```

Nice, here is something useful. There is some relationship between the two 
variables and it is coherent. The more one is in debt, the more expensive 
is to take more loans.

***

####Borrower Profile and Return: Monthly Income and Interest Rate

Talking about debt to income ratio, let's see the box plot of this variable 
related to Monthly Income. As I have seen before, there are big outliers here. 
Then, I am going to plot just income until $ 10,000 per month.

```{r, echo=FALSE, cache=TRUE, fig.width=9}
#set theme
theme_set(theme_minimal(14))

#box plot of income range by DI buckets
ggplot(aes(y = StatedMonthlyIncome, x = DebtToIncomeRatio.bucket), 
       data = subset(d,!is.na(ProsperRating..numeric.))) +
  coord_cartesian(ylim=c(0, 10000)) +
  geom_boxplot() +
  ggtitle("Distribution of Monthly Income by Debt to Income category\n")
```

The level of debt is directly related to income. It corroborates with what I 
have found before. If people with bigger Debt to Income Ratio pay higher 
interest rates and this ratio is negatively related to monthly income, then, 
the bigger income, the smaller the interest rate.

***

####Borrower Profile and Return: Homeowners and Interest Rate

If there is this relation with income, maybe there is some relationship 
with the assets of the borrower. Let me check the interest rate distribution 
when people are home owners or not. I will filter out borrower rate smaller than
39%

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#plot boxplot of borrowers rate diveded by IsHomeOwner
ggplot(aes(x = IsBorrowerHomeowner, y =BorrowerRate),
       data = subset(d,BorrowerRate<0.39 & !is.na(DebtToIncomeRatio) )) +
  scale_color_brewer(type = 'div') +
  geom_boxplot(color = "black") +
  ggtitle('BorrowerRate when Borrower is Home Owner')
  
```

Here I also found some relationship. I guess It couldn't be different . When you 
talk about risk, you talk about uncertainty. If you are confident that the 
borrower will pay the loan back, you can charge a lower interest rate. It is how
things work between countries, makes sense that it works in the same 
way between people. If someone earns more money per month or he has a house, you 
might be more confident that he will have conditions to pay you back.

***

####Risk and Loan Meta Data: Risk Scores and Defaulted Loans

Well...I am talking over risk and uncertainty....but, first of all, loans 
classified as risky really are riskier? Let's check the default ratio between 
risk scores.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#grouping by prosper rating, summarizing Risk category with counts of defaulted
#loans and total count and using it to calcultae the defaulted percentual
risk_groups <- group_by(filtered_d ,ProsperRating..numeric.)
d.risk_groups <- summarise(risk_groups,
                             Count = n(),
                             Defaulted = 
                               sum(LoanStatus.renamed=="Chargedoff/Defaulted"))

defaulted_by_risk <- mutate(d.risk_groups,
                            DefaultedPercent = Defaulted/Count,
                            notDefaulted= 1 - DefaultedPercent)

data2 <- melt(defaulted_by_risk[,c('notDefaulted','DefaultedPercent')])
row <- as.factor(rep(1:7,length=nrow(data2)))
df <- cbind(data2, row)

#plot a stacked bar chart showing the participation of defaulted loans by risk 
#category
ggplot(df, aes(x=row, y=value, fill=variable)) + 
  theme(legend.position = "none") +
  geom_bar(stat="identity") +
  xlab("\nProsper Rating") +
  ylab("Percentage\n") +
  scale_fill_economist()+   
  ggtitle("Proportion of defaulted loans \non each Risk Category")
  
```

Yes, they are. The number of borrowers that didn't pay the loan back are greater
at worst risk scores. 

***

####Risk, Return and Loan Meta Data: Are They All Related?

Ok... I want to check one last thing. Let's see the distribution of borrower 
rate by month. In order to put into perspective, I am also going to plot the 
default ratio by month and the total number of loans by month.

```{r, echo=FALSE, cache=TRUE, fig.height=9}
#set theme
theme_set(theme_minimal(14))

#grouping by the moth of origination, summarizing it with counts of defaulted
#loans and total count and using it to calcultae the defaulted percentual by 
#month
originDate_groups <- group_by(filtered_d ,LoanOriginationDate.month)
d.originDate_groups <- summarise(originDate_groups,
                             Count = n(),
                             Defaulted = 
                               sum(LoanStatus.renamed=="Chargedoff/Defaulted"))

defaulted_by_month <- mutate(d.originDate_groups,
                            DefaultedPercent = Defaulted/Count,
                            notDefaulted= 1 - DefaultedPercent)

l_months <- c('Jan','Feb','Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep','Oct',
              'Nov', 'Dec')

data2 <- melt(defaulted_by_month[,c('DefaultedPercent')])
row <- factor(rep(l_months,length=12),l_months)
df <- cbind(data2, row)


#plot number of loan by month
p1 <- ggplot(filtered_d, aes(x=LoanOriginationDate.month)) + 
  geom_bar() +
  ggtitle("# of loans by month")

#bar chart of defaulted ratio by origination month
p2 <- ggplot(df, aes(x=row, y=value, fill=variable)) + 
  theme(legend.position = "None") +
  geom_bar(stat = "identity") +
  xlab("\nOrigination month") +
  ylab("Percentage\n") +
  scale_fill_economist()+   
  ggtitle("Percentage of defaulted loans \nby origination month")

#plot the boxplot of borrower rate by origination month
p3 <- ggplot(filtered_d, aes(x=LoanOriginationDate.month, y=BorrowerRate)) + 
  geom_boxplot() + 
  ggtitle("Distribution of Borrower Rate \nby origination month")

#plot p1, p2 and p3 in the same figure
grid.arrange(p1,p2,p3, ncol=1)
  
```

Look to that. As the number of loans by month increases, the median of interest 
rate decreases, as well as the percentage Defaulted ratio. Right now, I can't 
say why , although I can think in a reason for this behaviour. For instance, in 
finance, if you want to reduce the risk of your investments, you should 
<a href="http://en.wikipedia.org/wiki/Modern_portfolio_theory#Diversification">
diversify</a> them. If I suppose that the interest rate is some function of the 
defaulted ratio, and then I diversified to whom I  lend money, I would expect 
that the defaulted ratio went down and then I could reduce the interest rate.

***

####Wrapping Up the Bivariate Section

In this section, I focused on understanding what makes someone be classified as 
risky. I have explored many variables. Many of them presented some correlation 
between what I was looking for, but not strong enough to allow me to formulate 
some hypotheses.

I have learned that loans with worst risk scores are, in fact, risky. They 
present a bigger defaulted ratio. As expected, I also have found that better 
risk classification is related to smaller interest rates and the amount that 
someone borrows.

Also, I verified that the more deep in debt someone is, related to his income, 
the more expensive is to take loans. My analysis has shown that the level of 
debt from someone is related to his level of income. I also have saw that when 
someone has a house, the average interest rate is smaller.

In the next section, I will try to find out how the term variable relates to 
what I already have found.


###Multivariate Section
<p>&nbsp;</p>
####Return,Profile and Meta Data: Interest Rate, Term and Debt Ratio

As I have seen in the univariate section, the dispersion of interest rate with
term of 36 month is greater than others. In order to understand that, I am going
to plot the Borrower rate by Term, adding some colors using the buckets that I
have created related to Credit to Debt Ratio.

```{r, echo=FALSE, cache=TRUE, fig.width=9}
#set theme
theme_set(theme_minimal(17))

#scatter plot of interest rate by DI ratio category, colored by term
ggplot(aes(x = DebtToIncomeRatio.bucket, y =BorrowerRate, color = Term.f ),
       data =  subset(filtered_d, !is.na(DebtToIncomeRatio.bucket))) +
  theme(legend.position = "bottom") +
  geom_point( position = position_jitter( h = 0)) +
  scale_color_economist() +  
  ggtitle('BorrowerRate by Term and Debt to Income Ratio') 
  
```

It seems that loans with term of 36 months are predominant at higher Debt to 
Income ratio. It doesn't necessarily mean something. 

***

####Borrower Profile and Meta Data: Percentages of D/I ratio Buckets by Term

Let me see how many loans are at at each bucket, divided by term. I am going to 
plot all buckets in one figure and just the worst debt to income buckets in 
another one.

```{r, echo=FALSE, cache=TRUE, fig.height=9, fig.width=9}
#set theme
theme_set(theme_minimal(17))

#include numbers in the place of buckets text
filtered_d$DebtToIncomeRatio.bucketN <- as.integer(
  filtered_d$DebtToIncomeRatio.bucket)

#grouping by the Term and DI bucket, then summarizing the groups to find out the
#percentage of each DI bucket on each term
term_groups <- group_by(subset(filtered_d, !is.na(DebtToIncomeRatio)),
                        Term.f,DebtToIncomeRatio.bucket)
d.di_by_term <- summarise(term_groups,n = n())
di_by_term <- mutate(d.di_by_term,
                     sum.n = sum(n),
                     "TotalPercentage" =  n/sum.n)
df <- di_by_term[,c("DebtToIncomeRatio.bucket",
                     "Term.f",
                     "TotalPercentage")]

#grouping and summarise higher DI ratio data
term_groups2 <- group_by(subset(filtered_d, !is.na(DebtToIncomeRatio) &
                                  DebtToIncomeRatio>1 ),
                         Term.f,DebtToIncomeRatio.bucket)

d.di_by_term2 <- summarise(term_groups2,n = n())
di_by_term2 <- mutate(d.di_by_term2, sum.n = sum(n),
                      "TotalPercentage" =  n/sum.n)

df2 <- di_by_term2[,c("DebtToIncomeRatio.bucket",
                      "Term.f",
                      "TotalPercentage")]

#define a array of colors to be uses in the chart
my_color_pallete <- c("#ece7f2", "#d0d1e6", "#a6bddb", "#74a9cf", "#3690c0", 
                      "#0570b0", "#045a8d", "#023858")


#plot a stacked bar chart of the percebtage of each bucket on Term category
p1 <- ggplot(df, aes(x=Term.f, y=TotalPercentage, 
                     fill=DebtToIncomeRatio.bucket)) + 
  geom_bar(stat="identity") +
  xlab("\nTerm") +
  ylab("Percentage\n") +
  scale_fill_manual(values = my_color_pallete ) +
  ggtitle("% of Debt to Income Bucket \non each Term")

#the same as before, but plotting just the higer DI buckets
p2 <- ggplot(df2, aes(x=Term.f, y=TotalPercentage, 
                      fill=DebtToIncomeRatio.bucket)) + 
  geom_bar(stat="identity") +
  xlab("\nTerm") +
  ylab("Percentage\n") +
  scale_fill_manual(values = my_color_pallete[5:8] ) +
  ggtitle("% of Debt to Income Bucket \non each Term (DI>=1)")

#plot p1 and p2 in one figure
grid.arrange(p1,p2, ncol=1)

```

I can see that all Terms have a small number of loans in the worst Debt to 
Income (D/I) buckets (equal to or greater than 1). On the other hand, when I 
filter out the smaller buckets, I see that the Term of 36 month has a bigger 
participation of buckets greater than 2 D/I ratio where the average borrowing 
rate is higher, as I already have seen before.

***

####Meta Data and Return: How Interest Rates with different Terms Changed Over Years?

I can't say that that is related to the dispersion of loans with Terms of 36 
months, but still I can't discard this hypothesis. I am going to plot the 
boxplot of the Borrower rate by Term and filtering out the D/I greater than 
0.14. I am also going to facet by year to see if the behaviour changes over 
time.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(17))

#plot boxplot of BorrowerRate by term and faceting by year
ggplot(aes(x = Term.f, y = BorrowerRate), 
       data = subset(filtered_d, DebtToIncomeRatio < 0.14)) +
  geom_boxplot() +
  facet_wrap(~LoanOriginationDate.year) + 
  ggtitle("Borrower Rate by Term and Year\n(D/I < 0.14)")

```

Interesting. The dispersion is not related to the D/I ratio, given that there is
dispersion even in Debt To Income Ratios smaller than 0.14. It remains true at 
almost every year where the Prosper Score were set (after 2008). Just in 2014 I
can't verified this relation.

***

####Meta Data and Borrower Profile: How D/I of Different Terms Changed Over Years?

Let me plot the defaulted ratio by Term to see if it sheds some light on what 
makes the 36 months more dispersed.I am going to faceted the next chart by year 
of ClosedDate and filter the loans with D/I rate smaller than 0.14 because it 
accounts for 55% of the overall data. 

```{r new_14, echo=FALSE, cache=TRUE, fig.width=9}
#set theme
theme_set(theme_minimal(14))

#update filtered_d
d$ClosedDate <- as.Date(d$ClosedDate,format = "%Y-%m-%d")
d$ClosedDate.year <- format(d$ClosedDate, "%Y")
filtered_d <- filter_data(d)

#grouping by the Term and Closed Year, then summarizing the grouped data to 
#find out the defaulted ration on each year
term_groups_14 <- group_by(subset(filtered_d, 
                                  DebtToIncomeRatio < 0.14 & 
                                    !is.na(ClosedDate.year)) ,
                              Term.f, ClosedDate.year)

d.term_groups_14 <- summarise(term_groups_14,
                             Count = n(),
                             Defaulted = 
                               sum(LoanStatus.renamed=="Chargedoff/Defaulted"))

defaulted_by_term_14 <- mutate(d.term_groups_14,
                            DefaultedPercent = Defaulted/Count)

#plot summaries as a scatter plot with  line chart to help visualize different 
#Terms
s_t <- "% of  Defaulted Loans \nby Closed Year(D/I < 0.14)"
ggplot(aes(x=ClosedDate.year, 
           y=DefaultedPercent,
           group = Term.f),
             data = defaulted_by_term_14) +
  geom_line(aes(color = Term.f)) + 
  geom_point(aes(color = Term.f)) + 
  scale_colour_brewer("Blues") +   
  ggtitle(s_t)

```

I can see that the loans with 12 month of Term, that also is the one with 
smaller Borrower Rate, presented a much smaller defaulted rate overall.

This chart also shows that the loans with Term of 60 months presented a smaller 
defaulted ratio than the loans with Term of 36 months in 2011 and in 2012. It 
just changed in 2013. 

As the default ratio of Terms of 60 months was becoming higher, the borrower 
rate was growing. Finally, borrowers rates with this term sharply increased in 
2014 after 2 years of higher defaulted ratio than the ratio of shorter terms.

It would explain why the borrower rate of loans with terms of 36 months are more
dispersed than those with 60 months. For me, it is due to the loans with longer 
terms have shown a smaller defaulted  ratio in early stages, accumulating data 
with smaller interest rates. It suggests  that I can't use the some of features 
without considering the date they are related to.

***

####Wrapping Up the Multivariate Section

In this section, I learned that the worst debt to Income Ratios are more common 
in loans with terms of 36 months. I also verified that the defaulted ratio 
changed differently over time when considered by Term, what may have affected 
the borrowers rates.


##Final Plots and Summary
<p>&nbsp;</p>

####First Figure: Defaulted Loans and Interest Rate by Month of Origination

The figure bellow shows how many loans were classified as defaulted on each 
month, as percentage, considering the date that they were originated. Also shows
how the borrower rates were distributed on each month. I called as defaulted 
every loan that was classified as defaulted or charged-off.

```{r, echo=FALSE, cache=TRUE, fig.height=9}
#set theme
theme_set(theme_minimal(14))

#grouping by Origination month, then summarizing the grouped data to 
#find out the defaulted ration on each month
originDate_groups <- group_by(filtered_d ,LoanOriginationDate.month)
d.originDate_groups <- summarise(originDate_groups,
                             Count = n(),
                             Defaulted = 
                               sum(LoanStatus.renamed=="Chargedoff/Defaulted"))
defaulted_by_month <- mutate(d.originDate_groups,
                            DefaultedPercent = Defaulted/Count,
                            notDefaulted= 1 - DefaultedPercent)

l_months <- c('Jan','Feb','Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep','Oct',
              'Nov', 'Dec')

data2 <- melt(defaulted_by_month[,c('DefaultedPercent')])
row <- factor(rep(l_months,length=12),l_months)
df <- cbind(data2, row)

#bar chart with defaulted ratio by month
p2 <- ggplot(df, aes(x=row, y=value, fill=variable)) + 
  theme(legend.position = "None") +
  geom_bar(stat = "identity") +
  scale_fill_economist()+   
  ggtitle("Percentage of Defaulted Loans \nBy Month of Origination")+ 
  ylab("Percentage\n") + 
  xlab("\nMonth")

#plot the boxplot of borrower rate by month, ,I am plotting it to help explain 
#my thoughts
p3 <- ggplot(filtered_d, aes(x=LoanOriginationDate.month, y=BorrowerRate)) + 
  geom_boxplot() + 
  ggtitle("Distribution of Borrower Rate \nBy Month of Origination")+ 
  ylab("Borrower Rate\n") + 
  xlab("\nMonth")

#plot p1 and p2 in one figure
grid.arrange(p2,p3, ncol=1)
  
#by(filtered_d$BorrowerRate,filtered_d$LoanOriginationDate.month,summary)
```

The percentage of defaulted loans increased from February (6%) to March (9%) with 
a peak in April (12%). The average borrower rate also increased during these 
months from 15% to 17% to 20%. Overall, the interest rate ranged from around 5%
to 35% in the data selected. Also, the sazonality in both features is similar. In
January, February and December were the smallest medians of Borrower rates 
(around 17%) and percentage of defaulted loans (5%, 6% and 6% again).

In other words, ss the percentage of defaulted loans increases, the average 
borrower rate also increases. With this chart, I suppose that both variable were
linked somewhat and this insight drove to my last analyzes on this data set.

***

####Second Figure: Borrower Rates By Term and Year of Origination

The next one, I choose It shows boxplots of the borrower rate splitted by term, 
showing separately the dispersion on each year. 

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(14))

#plot boxplot of Borrower rate by Term, faceting by year
s<-"Borrower Rate by Term in months\nand Year of Origination (D/I < 0.14)\n"
ggplot(aes(x = Term.f, y = BorrowerRate), 
       data = subset(filtered_d, DebtToIncomeRatio < 0.14)) +
  geom_boxplot() +
  facet_wrap(~LoanOriginationDate.year) + 
  ggtitle(s) + 
  ylab("Borrower Rate\n") + 
  xlab("\nTerm")

# for (i in seq(2010, 2014, 1)) {
# print(median(subset(filtered_d, Term==12 & 
#                       DebtToIncomeRatio < 0.14 & 
#                       LoanOriginationDate.year==i)$BorrowerRate))
# }

```

The average borrower rate of loans with term of 36 months and Debt to Income 
Ratio smaller than 0.14, increased from 2010 (19%) to 2012 (20%) and decreased 
to 2014 (12%), the smallest average rate for this term. The loans with term of 
60 months increased from 2010 (14%) to 2012 (21%) and decreased to 2014 (16%). 
The average borrower rate of terms of 60 months became higher than the term of 
36 month  since 2012. Although the interest rate for terms of 12 months has 
become more  dispersed, the maximum value for its average was around 12%.

Also, the figure above shows that the loans with 12 and 60 months have begun to 
be made in 2010, with the average borrower rate much smaller than loans with 36 
month. As pointed before, in 2011, the loans with the two Terms have already 
begun to be more dispersed, although the distance between the second and the 
third quartile always was more wide in the loans with Terms of 36 months.

I would say that the borrower rate of loans with term of 60 month has became 
really greater than the 36 months in 2014, when even the second quartile of the 
60  months almost became greater than the third quartile of loans with term of 
36  months.

This chart helped me understand how the dynamic of the borrowers rates changed 
through the years. Initially I had observed that loans with 36 month had been 
more dispersed than the 60 month and I assumed that it was a characteristic of 
the loans. After this chart, I wondered if there were another reason for the 
dispersion.

***

####Third Figure: Defaulted Loans By Term and Closing Year 

Then, following my previous reasoning, I plotted the ratio of the defaulted loans
by the year when the loans were closed, splitting them by term.

```{r, echo=FALSE, cache=TRUE}
#set theme
theme_set(theme_minimal(14))

#grouping by the Term and Closed Year, then summarizing the grouped data to 
#find out the defaulted ration on each year
filtered_d_v2 <- subset(filtered_d, DebtToIncomeRatio < 0.14 &
                                    !is.na(ClosedDate.year))
term_groups_14 <- group_by( filtered_d_v2,Term.f, ClosedDate.year)

d.term_groups_14 <- summarise(term_groups_14,
                             Count = n(),
                             Defaulted = 
                               sum(LoanStatus.renamed=="Chargedoff/Defaulted"))

defaulted_by_term_14 <- mutate(d.term_groups_14,
                            DefaultedPercent = Defaulted/Count)

term_groups_org <- group_by( filtered_d_v2,Term.f, LoanOriginationDate.year)

d.term_groups_org <- summarise(term_groups_org,
                             Interest_Median = median(BorrowerRate))

#plot summaries
s<-"% of defaulted loans by\n Closed Year (D/I < 0.14)"
s2<-"Median(Borrower Rate) by \n Origination Year (D/I < 0.14)"

#plot the defaulted loans percentage by closed year and Term
p1 <- ggplot(aes(x=ClosedDate.year, 
           y=DefaultedPercent,
           fill = Term.f),
             data = defaulted_by_term_14) +
  theme(legend.position = "bottom") +
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_economist(name = "Term \nin Months") + 
  ggtitle(s) + 
  ylab("Percentual\n") + 
  xlab("\nClosed Year")

#plot the interest rate median by origination year and Term
p2 <- ggplot(aes(x=LoanOriginationDate.year, 
           y=Interest_Median,
           fill = Term.f),
             data = d.term_groups_org) +
  theme(legend.position = "bottom") +
  geom_bar(stat="identity", position = "dodge") + 
  scale_fill_economist(name = "Term \nin Months") + 
  ggtitle(s2) + 
  ylab("Median(Borrower Rate)\n") + 
  xlab("\nOrigination Year")

#plot p1 and p2 in one figure
grid.arrange(p1, ncol=1)

```

The percentage of defaulted loans with term of 60 months increased from 2012 
(20%) to 2013 (33%) and decreased in 2014 (29%). In comparison, the defaulted 
ratio with term of 36 month has become smaller than the Term of 60 months when 
it slightly decreased from 2012 (23%) to 2013 (22%) and in 2014 (18%). The 
defaulted ratio of loans with 12 months never rose from 7%.

In the chart above, I decided to use the year when the loans were closed 
because I had imagined that years when were verified a increase in defaulted 
ratio, the borrower rates probably would also increase.

As can be seen, in fact, it happened. In 2013 and 2014, when the defaulted ratio
of the loans with term of 60 month sharply increasesd(as shown above), the 
median of borrowers rate also increased (as shows the Second Figure).

I choose this last figure because this relation between defaulted ratio and 
borrower rate helped me to understand why the borrower rate changed his 
behaviour throughout the years between different terms.


##Reflection

At the beginning, I didn't know what to looking for in the data set. I believe 
that the number of features confused me at first and I spent some time trying to
understand the variables before start the project. Then, I just started to poke 
around the data, trying to find something interesting to explore, without 
success. Then, I decided to make some assumptions and try to check them in the 
collection. After that, my project took off.

As I am used to Scilab and Pandas, at the beginning I was thinking that the 
syntax of R would not be an issue for me. I wasn't totally right. I suffered 
trying to group and reshape the data set, but after that I understood the basic 
logic of the libraries I was using, it wasn't so painful.

During my analysis, I verified that many variables that I have believed that 
would have a greater effect on Borrowers rate, actually presented just a smaller
role. Maybe if I tried to build a model to predict the borrower rate, all these 
variables would be useful, but not alone, just together. 

At the end, I just found two variable that helped me to understand the borrower 
rate variations. Debt to Income Ratio and the overall Defaulted ratio. The last 
one was curious. It seems be connected to the level of all interest rates for a 
particular year.

I also did not find any good reason to transform any variable. I have expected 
to find something, given that the data set is related to money.

I stopped my analysis when I have started to explore how the distributions of 
this collection had been changing over year. Maybe if I crossed this data set 
with economic indicators I could find some interesting stories. For instance, 
would be interesting to check if there is a correlation between unemployment 
rate and defaulted ratio. It probably would affect the overall borrower rates.
