<h1><b>Understanding Prosper Loan Data Set</b></h1>
<p><small><em>Uirá Caiado. April 7, 2015</em></small></p>
------------------
```{r, include=FALSE, cache=FALSE}
library("plyr")
library("tidyr")
library("dplyr")
library(ggplot2)
library("ggthemes")
library(gridExtra)
library(reshape2)

theme_set(theme_minimal(24))


#rm(list = setdiff(ls(), lsf.str()))
```

<h2>Abstract</h2>

bla bla

<h2>Introduction</h2>

I have to tell the truth. Initially, the number of variables intimidate. There are 113,937 rows in this data set and 81 columns (variables). Besides that, I still do not understand the origin of those data very well. I have searched a little and this data seems to come from Prosper, a <em><a href="http://en.wikipedia.org/wiki/Peer-to-peer_lending">peer-to-peer lending</a></em> marketplace. I have already seem a brazilian website start to do something similar here in my country, Brazil, but our "SEC" shut them off. Just finantial companies can lending money here.  

Well, I am not sure what I can do with this data set and would like to approach the most important variables, but I do not know yet how to choose. Reading the docs, I found out that there are some variables that is just applicable for loans starting on 2009. First, let's check how many of the data points are in each year and how much money is involved.

```{r, include=FALSE, cache=FALSE}
#data <- read.csv(s_fname)

#str(data)
#summary(data)
```


```{r}

s_fname <- "~/git/ProsperLoan_EDA/prosperLoanData.csv"

data <- read.csv(s_fname)

#creating new variables
l_months <- c('Jan','Feb','Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep','Oct',
              'Nov', 'Dec')

data$LoanOriginationDate <- as.Date(data$LoanOriginationDate,format = "%Y-%m-%d")
data$LoanOriginationDate.year=format(data$LoanOriginationDate, "%Y")
data$LoanOriginationDate.month=format(data$LoanOriginationDate, "%b")
data$LoanOriginationDate.month <- factor(data$LoanOriginationDate.month, l_months)


p1 <- ggplot(aes(x = LoanOriginationDate.year), data = data) +
  geom_histogram() +
  ggtitle('Year')

p2 <- ggplot(aes(x = LoanOriginationDate.year, y = LoanOriginalAmount), data = data) +
  geom_bar(stat="identity") +
  ggtitle('Year')


p3 <- ggplot(aes(x = LoanOriginationDate.month), data = data) +
  geom_histogram() +
  ggtitle('Month')

p4 <- ggplot(aes(x = LoanOriginationDate.month, y = LoanOriginalAmount), data = data) +
  geom_bar(stat="identity") +
  ggtitle('Month')


grid.arrange(p1, p3, ncol=1)

grid.arrange(p2, p4, ncol=1)


```

It actualy makes sense, as Prosper also had <a href = "http://en.wikipedia.org/wiki/Prosper_Marketplace#Loan_performance_prior_to_July_2009">some issues</a> with the SEC and have to close the site from 2008 untill July 2009. 


```{r, echo=FALSE}
plot(cars)
```

<h2>Analysis</h2>

<h3>exploring 1</h3>

bla

<h3>plotting 2</h3>

bla

<h3>talking 3</h3>

bla


<h2>Final Plots and Summary</h2>

bla

<h2>Reflection</h2>

bla

